Explicit Scale Simulation for analysis of RNA-sequencing  with ALDEx2

### 2025-03-11 go3.Rmd
submitted version Dec 6

---
title: "Explicit Scale Simulation for analysis  of RNA-sequencing  with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document 
header-includes:
  - \usepackage{setspace}
package: ALDEx2
abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total mRNA expression) of the underlying biological systems. Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses. We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates. We demonstrated that these errors can be mitigated by accounting for uncertainty about scale using a *scale model*, which we integrated into the ALDEx2 R package. This article provides new insights into those methods, focusing on the application to transcriptomic analysis. Here we provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses. We show that these scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance. We demonstrate the utility of that scale models   built based on known housekeeping genes in complex metatranscriptomic datasets. Thus this work provides example and practical guidance on how to incorporate scale into transcriptomic analysis.

@@

\doublespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a 
compositional sampling step as only a fixed-size subsample of nucleic acid is 
carried over to subsequent steps. 
Thus, with
each sampling step the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost. In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process[@lovell:2011;@Quinn:2019aa;@nixon2023scale]. 
Increasingly, researchers are turning to modified experimental protocols
 in an attempt to recover the lost
biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
However, these protocols only recover scale variation at
the intermediate step in the sample preparation protocol where the intervention was made. 
In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an open challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Li:2022aa], 
  exhibited as dataset-dependent 
FDR control and by the disconnect between statistical and biological significance[@Schurch:2016aa]. In current practice,  these issues are addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and  a large difference between groups is used to identify interesting transcripts or genes for follow-up analysis[@Schurch:2016aa]. 
This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but
 is known to not   appropriately control the  FDR  
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing data where the mean change between groups is non-zero. Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other. 
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. 
The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@Li:2022aa;@nixon2023scale,@Nixon2024B].

The  three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.  Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).  Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2023scale,@Nixon2024B].

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret. Because of this different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives.

The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. We recently modified ALDEx2  to explicitly model the scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B]. Here, we briefly introduce review modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given in [@nixon2023scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate LFC significance.  At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets. First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013]. However, this step was modified recently to account for scale uncertainty and misspecification [@nixon2023scale,@Nixon2024B] via a scale model, explained  with more details in [@nixon2023scale;@Nixon2024B] and summarized in the next paragraph. Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation. These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework. For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013]. These are calculated on a \(\log_2\) scale. For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2023scale;@Nixon2024B].

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al. [@nixon2023scale] term  a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al. [@nixon2023scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.  They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate. With this modificaiton, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

@@

This formulation is quite flexible [@nixon2023scale;@Nixon2024B]. In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution that adds scale uncertainty.  The default approach controls only the degree of uncertainty in the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the advanced or 'informed' configuration, \(\mu\) takes different values for each group, and this controls the location of the LFC assumption. An example of both the default and informed approaches is given for a microbiome dataset in @Nixon2024B, and here we show that these approaches also work well in transcriptome and metatranscriptome datasets. These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.

@@

Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different. Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total.  Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others. Below we include an intuitive explanation of how incorporating scale uncertainty achieved this outcome using a setting of \(\gamma=0.5\), which is in line with the recommendations of [@Nixon2024B] and that gives results comparable to those proposed in [@Schurch:2016aa].

@@

```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))
# volcano
#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
#title('D: DESeq2 volcano', adj=0, line= 0.8)
#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
#  pch=19, cex=0.5)
#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
#  pch=19, cex=0.5)
#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')

@@

#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
#  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
#title('A: DESeq2 effect', adj=0, line= 0.8)
#points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
#  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
#  pch=19, cex=0.5)
#points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
#  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
#  pch=19, cex=0.5)
#abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')

@@

# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5, col=rgb(0,0,1,0.3), ylim=c(0,4.5), 
#   xlab='g=0 Dispersion', ylab='g=0.5 | g=1 Dispersion')
# points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5, col=rgb(1,0,0,0.3))
# abline(0,1, lty=2)
# title(main='B', line=-1.2, adj=0.5)
# legend(0,4, legend=c( "0.5", "1"), 
#   pch=19, col=c("red", "blue"))

@@

It is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa]. Note that there is considerable variation in recommended cutoff values[@Schurch:2016aa]. Here, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line with the recommendations of [@Schurch:2016aa]. the fold change limits are shown by the dashed grey lines in Figure 1. The  \(2^{1.4}\) fold change cutoff identifies a similar number of transcripts as does  ALDEx2 using \( \gamma = 0.5 \) which identifies `r length(sig.s.ald)`.  Supplementary Figure 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify those transcripts that are very sensitive to scale uncertainty. In this supplementary figure we see that even adding a very small amount of scale \( \gamma = 0.1 \) reduces the number of significant transcripts by more than half. This allows us to those low-dispertion transcripts that were  significant only because of an absence of scale uncertainty. In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs [@Nixon2024B].

>>>>

---
title: "Explicit Scale Simulation for analysis of RNA-sequencing  with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
  - name: Justin Silverman
    affiliation: 
      College of Information Sciences and Technology and Department of Medicine, Pennsylvania   State University
      Department of Medicine, Pennsylvania State University
      Institute for Computational and Data Science, Pennsylvania State University
      Department of Statistics, Pennsylvania State University
cite-method: citeproc
citeproc: TRUE
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
package: ALDEx2

abstract: 
  In high-throughput sequencing (HTS) studies,  sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system. Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses. We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates. We demonstrated that these errors can be mitigated by accounting for uncertainty using a *scale model*, which we integrated into the ALDEx2 R package. This article provides new insights focusing on the application to transcriptomic analysis. We provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses. These scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance. We demonstrate the utility of scale models built based on known housekeeping genes in complex metatranscriptomic datasets. Thus this work provides guidance on how to incorporate scale into transcriptomic data sets.

@@

\doublespacing 
\singlespacing

High-throughput sequencing (HTS) is a ubiquitous
tool used to explore many biological phenomenon such as gene expression
(single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial
community composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a
sample from the environment, making a library, multiplexing (merging) multiple
libraries together, and then applying a sample of the multiplexed library to a
flow cell. Each of these steps is a compositional sampling step as only a
fixed-size subsample of nucleic acid is carried over to subsequent steps. Thus,
with each sampling step the connection between the actual size of the sampled
DNA pool and the scale (e.g., size, microbial load, or total gene expression) of
the measured biological system is degraded or lost. In the end, the information
contained in the data relates only to relative abundances and  has an arbitrary
scale imposed by the sequencing
process [@lovell:2011;@Quinn:2019aa;@nixon2024scale]. Researchers
can use modified experimental protocols [@Jiang:2011aa;@Loven:2012aa;@Vandeputte:2017aa;@Props:2017aa] or machine learning methods [@Nishijima:2024aa] to uncover the
biological variation in scale. However,  wet-lab protocols only provide information on
the size of the data downstream of the step in the sample preparation protocol where the
intervention was made and  introduce an additional source of variation that must be accounted for [@Loven:2012aa]. The computational methods developed for microbiome analysis have low correlation with the actual scale but are useful [@Nishijima:2024aa]. In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an open
challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue is
poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Ge:2021aa;@Li:2022aa], exhibited
as dataset-dependent FDR control and by the disconnect between statistical and
biological significance [@Schurch:2016aa]. In current practice,  these issues are
addressed  by a dual-filtering method, whereby both a low p-value (or
equivalently a low q-value following FDR correction [@storey2003positive]) and a
large difference between groups is used to identify interesting transcripts or
genes for follow-up analysis [@Cui:2003aa;@Schurch:2016aa]. This double-filtering approach is
graphically exemplified by the volcano plot [@Cui:2003aa], but is known to not
appropriately control the  FDR [@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second
issue is poor performance when analyzing data where the mean change between
groups is non-zero [@nixon2024scale]. Such asymmetric data can arise when a gene set is expressed
in one group but not the other, or when one group contains different gene
content from the other. This type of data frequently arises in in-vitro
selection experiments (SELEX), transcriptome analysis, and microbiome analysis
[@Wu2021]. The third issue is that the actual scale of the environment is often
a major confounding variable during analysis [@nixon2024scale;@Nishijima:2024aa]. The final issue is that these problems become more pronounced as more
samples are collected; that is, more information results in a worsening of the
accuracy of the analysis [@Li:2022aa;@nixon2024scale;@Nixon2024B].

The  four problems were recently shown by Nixon et al. [@nixon2024scale]
to be a result of a mismatch between the underlying size or scale of the system
and the assumptions of the normalizations used for the analysis of HTS.
Biological variation in scale often represents an important unmeasured
confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the
cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA
content than  non-transformed cells [@Nie:2012aa], and this dramatically skews
transcriptome analysis unless spike-in approaches are used [@Loven:2012aa]. In addition, wild-type and mutant
strains of cell lines, yeast or bacteria  have different growth rates and RNA
contents under different conditions, which  affect our ability to identify truly
differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As
another example, the total bacterial load of the vaginal microbiome differs by
1-2 orders of magnitude in absolute abundance between the healthy and bacterial
vaginosis states [@Zozaya:2010], and the composition between these states is
dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of
any of these systems includes both relative change (composition) and absolute
abundance (scale).  Current methods access only the compositional information
yet make implicit assumptions about the scale [@nixon2024scale,@Nixon2024B].

Recently, Nixon et al. [-@nixon2024scale] showed that the challenge of
non-biological variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations make
some assumption about  scale but these implicit assumptions are often
inappropriate and difficult to interpret. This causes different normalizations
to provide different outputs when applied to the same dataset
[@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa].
Intuitively, normalizations in widespread use  assume that either all samples
have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM
[@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample
can be chosen as a reference to which the others are scaled e.g. the TMM
[@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982];
or that different sub-parts of each sample maintain a constant scale across
samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is
appropriate e.g the CLR [@Aitchison:1986] and its derivatives.

The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict
assumption about scale through the CLR normalization [@nixon2024scale]. This assumption was often close enough to the true value to be useful, but was not always the a good estimate and could be outperformed by other normalizations [@Yerke:2024aa].
Nixon et al. [@nixon2024scale] showed that better scale assumptions resulted in
more reproducible data analysis including better control of both false positive
and false negative results. We recently modified ALDEx2  to explicitly model the
scale over a range of reasonable normalization parameters, and showed
significant improvements in performance in microbiome and in-vitro selection
experiments [@Nixon2024B]. Here, we briefly  review these modifications and show how scale uncertainty can greatly improve modeling in transcriptome
and meta-transcriptome datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given in
[@nixon2024scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote
the \emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to
feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we
can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the
biological system from which sample \(n\) was obtained. We can think of
\(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\)
(e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e.,
proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note
that with these definitions \(\mathbf{W}\) can be written as the element-wise
combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the
logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
\mathbf{W}^{\perp}_{n}\).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose
toolbox to  model the uncertainty of HTS data and to use that model to estimate
the underlying LFC (log-fold change) significance.  At a high-level, ALDEx2  has
three connected components to estimate the uncertainty inherent in HTS datasets.
First, the tool accounts for the uncertainty of the sequencing counts using
Dirichlet multinomial sampling to build a probabilistic model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly,
ALDEx2 uses the centred log-ratio transformation to scale the data
[@fernandes:2013]. However, this step was modified recently to account for scale
uncertainty and misspecification [@nixon2024scale,@Nixon2024B] via a scale
model, explained  with more details in [@nixon2024scale;@Nixon2024B] and
summarized in the next paragraph. Finally, a standard null-hypothesis test and a
non-parametric estimate of mean standardized difference are used to report on
the finite sample variation. These sources of uncertainty and variation are
combined via reporting the expected values from a Monte-Carlo simulation
framework. For simplicity, we use the term 'difference' to refer to the absolute
difference between groups, and 'dispersion' to refer to the  within-condition
difference or pooled variance as defined in [@fernandes:2013]. These are
calculated on a \(\log_2\) scale. For more details on ALDEx2 see
[@fernandes:2013;@fernandes:2014;@nixon2024scale;@Nixon2024B].

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a
specialized type of statistical model which Nixon et al. [@nixon2024scale] term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al.
[@nixon2024scale] generalized the concept of normalizations by introducing  the
concept of a \textit{scale model} to account for potential error in the centred
log-ratio normalization step.  They did this by including a model for
\(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by ALDEx2 makes the
assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the
geometric mean of sample n, which while being a random variable, is essentially
constant across each Monte-Carlo replicate, but that differs between samples.
With this modification, ALDEx2 can be generalized by considering probability
models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean
\(1/G_{n}\). For example, the following scale model generalizes the CLR:

@@

This formulation is quite flexible [@nixon2024scale;@Nixon2024B]. In the simple
or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter
drawn from a log-Normal distribution[@nixon2024scale]. Adding scale uncertainty with the \(\gamma\) paramenter  controls only the degree of uncertainty of the CLR assumption for the
\(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and
\(x_{n}=0\) denotes control). In the advanced or 'informed' configuration,
\(\mu\) takes different values for each group and  controls the location of
the LFC assumption; combining \(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the location and the scale. An example of both the default and informed approaches is
given for a microbiome dataset in [@Nixon2024B] showing increased sensitivity and specificity. Here we show that these
approaches also work well in transcriptome and metatranscriptome datasets. These
modifications are  instantiated in ALDEx2 which is  the first software package
designed for SSRV-based inference.

@@

Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast
transcriptome dataset to compare gene expression between a wild-type strain and
a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several
RNA-seq tools for their power  to detect the set of differentially abundant
transcripts identified in the full dataset when the data was subset
[@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of
transcripts with different tools identifying between between 65% to >80% of all
transcripts as being significantly different. Since the majority of transcripts
were significantly different, the authors suggested that it was more appropriate
to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995]
corrected p-value (q-value) plus a difference cutoff  to limit the number of
identified transcripts to a much smaller fraction of the total.  Nixon et al,
@Nixon2024B showed that adding even a small amount of  scale uncertainty with
ALDEx2  dramatically reduced the number of significant transcripts identified,
removing the need for the dual cutoff approach in this dataset and others. Below
we include an intuitive explanation of why and how incorporating scale uncertainty
achieved this outcome using a setting of \(\gamma=0.5\). The approach is in line with
the recommendations of [@Nixon2024B] and  gives results comparable to those
proposed in [@Schurch:2016aa].

@@

```{r plot1, echo=FALSE, warning=FALSE, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))
# volcano plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300),
# col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
# title('D: DESeq2 volcano', adj=0, line= 0.8)
# points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des
# ] + 1e-300), col=rgb(1,.66,0,0.3), pch=19, cex=0.5)
# points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s
# .ald] + 1e-300), col=rgb(1,0,0,1), pch=19, cex=0.5) abline(v=c(-1.4,1.4),
# lty=2,lwd=2, col='grey')

@@

#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange,
#xlim=c(0,5), col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
#title('A: DESeq2 effect', adj=0, line= 0.8)
#points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
#res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), pch=19, cex=0.5)
#points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
#res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), pch=19, cex=0.5)
#abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')

@@

# plot(yst.all$diff.win, yst.1.all$diff.win, pch=19, cex=0.5,
# col=rgb(0,0,1,0.3), ylim=c(0,4.5), xlab='g=0 Dispersion', ylab='g=0.5 | g=1
# Dispersion') points(yst.all$diff.win, yst.s.all$diff.win, pch=19, cex=0.5,
# col=rgb(1,0,0,0.3)) abline(0,1, lty=2) title(main='B', line=-1.2, adj=0.5)
# legend(0,4, legend=c( "0.5", "1"), pch=19, col=c("red", "blue"))

@@

It is common practice to use a dual-cutoff by choosing transcripts based on a
thresholds for both q-values and fold-changes  [@Cui:2003aa;@Schurch:2016aa].
Note that there is considerable variation in recommended cutoff
values[@Schurch:2016aa]. Here, applying a dual-cutoff using a heuristic of at
least a \(2^{1.4}\) fold change  reduces the number of significant outputs to
193 for DESeq2 and to 186 for ALDEx2. This cutoff was chosen for convenience and
is in-line with the recommendations of [@Schurch:2016aa] with the fold change limits
 shown by the dashed grey lines in Figure 1. The  \(2^{1.4}\) fold change
cutoff identifies a similar number of transcripts as does  ALDEx2 using \(
\gamma = 0.5 \) which identifies `r length(sig.s.ald)`.  Supplementary Figures 1
and 2 shows how to use the \texttt{aldex.senAnalysis()} function to identify
those transcripts that are very sensitive to scale uncertainty. In this
supplementary figure we see that even adding a very small amount of scale \(
\gamma = 0.1 \) reduces the number of significant transcripts by more than half.
This allows us to ignore those low-dispersion transcripts that were  significant only
because of an absence of scale uncertainty. In practice, we suggest that a
\texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental
designs [@Nixon2024B].
